---
title: "Introduction to sovpal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sovpal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 2
)
```

```{r setup}
library(sovpal)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## Introduction

The `sovpal` package provides color palettes derived from Soviet industrial
standards and military paint specifications. The palettes span qualitative,
sequential, and diverging types, with full ggplot2 integration.

### Design Philosophy

The standards-based palettes (`gost14202`, `gost_signal`, `soviet_military`,
`steppe`, `hazard`) carry fixed hex values traceable to specific documents —
GOST paint card numbers, paint manufacturer references, or military paint
standards. Modifying these values would alter the historical record.

The `constructivist` palette is evocative rather than standards-derived; its
colors are not fixed in the same sense.

Palettes come in two tiers. *Archival* palettes are faithful to their source.
*Visualization-optimized* palettes (suffixed `_lines`, `_core`, `_cvd`) select
or reorder colors that already appear in an archival palette for distinctness,
white-background contrast, or colorblind safety — they never introduce a new
hex value. To slice or subset further, index the returned named vector with
base R, e.g. `sovpal("gost14202")[c("water", "fire")]`.

## Installation

```{r, eval = FALSE}
remotes::install_github("kinelhu/sovpal")
```

## Available Palettes

```{r all-palettes, fig.height = 6}
show_all_palettes()
```

### Individual Palettes

**`gost14202` — GOST 14202-69 Industrial Pipeline Identification**

```{r pal-gost14202}
show_palette("gost14202")
```

**`soviet_military` — Soviet WWII and Cold War Vehicle/Aircraft Colors**

```{r pal-soviet-military}
show_palette("soviet_military")
```

**`constructivist` — Soviet Constructivist Graphic Design**

```{r pal-constructivist}
show_palette("constructivist")
```

**`gost_signal` — GOST 12.4.026 Mandatory Safety Signal Colors**

```{r pal-gost-signal}
show_palette("gost_signal")
```

**`steppe` — Soviet Camouflage Field Colors (Sequential)**

```{r pal-steppe}
show_palette("steppe")
```

**`hazard` — Safe to Danger (Diverging)**

```{r pal-hazard}
show_palette("hazard")
```

### Visualization-optimized palettes

These reuse only canonical hex values from the archival palettes above.

```{r pal-viz}
show_palette("gost14202_lines")
show_palette("constructivist_core")
show_palette("hazard_cvd")
show_palette("hazard_warm")
```

## Basic Usage

```{r basic-retrieval}
sovpal("gost14202")

sovpal("gost14202", n = 4)
```

Interpolate a continuous ramp:

```{r continuous}
ramp <- sovpal("steppe", n = 100, type = "continuous")
length(ramp)
```

Inspect metadata with `palette_info()`:

```{r palette-info}
palette_info("gost14202")
```

```{r hazard-info}
info <- palette_info("hazard")
cat(info$recommended_use, "\n")
cat(info$cvd_note, "\n")
```

Base R usage:

```{r base-r-plot, fig.height = 3.5}
barplot(
  1:8,
  col       = sovpal("gost14202"),
  main      = "GOST 14202-69 Pipeline Identification Colors",
  names.arg = names(sovpal("gost14202")),
  las       = 2,
  cex.names = 0.75
)
```

## White-Background and Colorblind-Safe Usage

Rather than altering colors at runtime, sovpal ships explicit
visualization-optimized palettes that reuse only canonical hex values. Request
them by name.

### Color scales: pick a lines palette

For points and lines on white, use the `_lines` palette. The light yellow and
dark brown stops of `gost14202` are dropped — and the choice is visible in the
call, not hidden:

```{r lines-demo, eval = has_ggplot2}
library(ggplot2)

ggplot(iris, aes(Sepal.Length, Sepal.Width, colour = Species)) +
  geom_point(size = 2) +
  scale_color_sovpal("gost14202_lines") +
  theme_minimal()
```

### Fill scales

`scale_fill_sovpal()` uses the named palette exactly as defined. For
Constructivist on white, use `constructivist_core`, which omits the
low-contrast cream ground:

```{r constructivist-fill, eval = has_ggplot2, fig.height = 3.5}
ggplot(iris, aes(Species, Sepal.Length, fill = Species)) +
  geom_boxplot() +
  scale_fill_sovpal("constructivist_core") +
  theme_minimal()
```

### Checking contrast

`palette_info()` reports each color's WCAG contrast ratio against white and
flags any below the 3:1 non-text threshold:

```{r contrast-check}
info <- palette_info("constructivist")
info$contrast_white
info$low_contrast_on_white
```

### Using `sovpal()` directly

`sovpal()` returns a plain named vector, so base-R indexing covers any custom
selection or slice:

```{r direct-subset}
# Select colors by name
sovpal("gost14202")[c("water", "fire", "air", "toxic", "alkali", "other")]

# Or just use the ready-made lines palette
sovpal("gost14202_lines")
```

## ggplot2 Integration

```{r ggplot2-bar, eval = has_ggplot2, fig.height = 3.5}
library(ggplot2)

df_fluid <- data.frame(
  fluid = factor(names(sovpal("gost14202")), levels = names(sovpal("gost14202"))),
  flow  = c(120, 95, 60, 80, 45, 70, 110, 55)
)

ggplot(df_fluid, aes(fluid, flow, fill = fluid)) +
  geom_col() +
  scale_fill_sovpal("gost14202") +
  labs(title = "GOST 14202-69: Pipeline Colors", x = NULL, y = "Flow rate (L/min)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r ggplot2-scatter, eval = has_ggplot2, fig.height = 3.5}
ggplot(iris, aes(Sepal.Length, Sepal.Width, colour = Species)) +
  geom_point(size = 2.5) +
  scale_color_sovpal("soviet_military") +
  labs(title = "soviet_military") +
  theme_minimal()
```

```{r ggplot2-heatmap, eval = has_ggplot2, fig.height = 3.5}
# Hazard diverging palette -- risk scores, hazard ratios, survival data
m <- outer(1:10, 1:10, function(x, y) (x - 5.5) * (y - 5.5))
df_hm <- data.frame(
  x = rep(1:10, 10),
  y = rep(1:10, each = 10),
  z = as.vector(m)
)
ggplot(df_hm, aes(x, y, fill = z)) +
  geom_tile() +
  scale_fill_sovpal("hazard", discrete = FALSE) +
  labs(title = "Hazard Palette -- Diverging Risk Heatmap", fill = "Score") +
  theme_minimal() +
  coord_equal()
```

The diverging palettes use a light cream midpoint that reads as a zero-point and
makes both arms pop. `hazard` itself keeps its archival GOST green/grey/red. For
plots, `hazard_warm` (green/cream/red) is the most legible, and `hazard_cvd`
(blue/cream/red) carries the same safe-to-danger semantics with colorblind-safe
hues for red-green-deficient audiences:

```{r ggplot2-heatmap-cvd, eval = has_ggplot2, fig.height = 3.5}
ggplot(df_hm, aes(x, y, fill = z)) +
  geom_tile() +
  scale_fill_sovpal("hazard_cvd", discrete = FALSE) +
  labs(title = "hazard_cvd -- Colorblind-Safe Diverging", fill = "Score") +
  theme_minimal() +
  coord_equal()
```

```{r ggplot2-sequential, eval = has_ggplot2, fig.height = 3.5}
# Steppe sequential gradient
df_seq <- data.frame(x = factor(1:8), y = c(3, 7, 2, 9, 5, 6, 4, 8))
ggplot(df_seq, aes(x, y, fill = y)) +
  geom_col() +
  scale_fill_sovpal("steppe", discrete = FALSE) +
  labs(title = "Steppe Sequential Palette") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Historical Notes

### GOST 14202-69 — Pipeline Identification

GOST 14202-69 (*Truboprovody promyshlennykh predpriyatiy. Opoznavatelnaya
okraska, preduprezhdayushchie znaki i markirovochnye shchitki*) was adopted in
1969 and standardized the color-coding of industrial pipelines across all
Soviet enterprises. The standard specified eight base colors corresponding to
fluid categories, referenced by paint card number. The `gost14202` hex values
correspond to those card-indexed samples. The system remains mandatory in
post-Soviet industrial facilities operating under inherited Soviet technical
regulations.

### GOST 12.4.026 — Safety Signage

GOST 12.4.026 standardized workplace safety signage colors across the Soviet
Union. The four signal colors encode hierarchical urgency: red (immediate
danger, prohibition), yellow (caution, warning zones), green (safe conditions,
evacuation routes), and blue (mandatory action, PPE requirements). Post-Soviet
revisions preserved all four signal colors with minimal change.

### 4BO Protective Green

4BO (*Zashchitnyy*) is the canonical Soviet military vehicle exterior color,
adopted in the 1930s and used continuously through WWII and the Cold War. The
hex value #51653F is based on Vallejo Model Color 70.609 and corroborated by AK
Interactive spectrographic measurements of surviving vehicle examples. It
appears in `soviet_military` (`green_4bo`) and `steppe` (stop 2).

### Cockpit Turquoise

Soviet aircraft cockpits from the 1940s through the 1970s were finished in a
distinctive grey-turquoise used across diverse airframes including the MiG-15,
MiG-21, Tu-95, and Su-27. The value #5B8F8C is derived from AK Real Colors
RC206 (*Interior Turquoise Green*), formulated from spectrographic analysis of
preserved cockpit interiors.

### Constructivist Poster Design

Soviet Constructivism (1920s--1930s) developed a visual language centered on
a small set of high-contrast colors: red, black, and white, with yellow and
blue as secondary elements. The `constructivist` palette is evocative of this
corpus rather than derived from a specific technical standard. The cream
(#E8DFC8) reflects the typical ground color of printed poster reproductions
and is low-contrast on white backgrounds.
