---
title: "Introduction to sovpal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sovpal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 3.6,
  dpi = 110
)
```

```{r setup}
library(sovpal)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r theme, include = FALSE}
# An opinionated visual language for the showcase. theme_linedraw evokes the
# ruled, boxed look of Soviet technical drawing (the GOST/ESKD drafting world
# these colors come from); we lighten its grid so the COLOR carries the plot.
if (has_ggplot2) {
  library(ggplot2)
  theme_sovpal <- function(base_size = 11) {
    theme_linedraw(base_size = base_size) +
      theme(
        panel.grid.major = element_line(linewidth = 0.2, colour = "grey80"),
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey30"),
        legend.key       = element_rect(colour = NA)
      )
  }
  theme_set(theme_sovpal())
}
```

`sovpal` provides color palettes drawn from the Soviet industrial world — GOST
pipeline-identification standards, military paint specifications, safety
signage, and Constructivist graphic design. The colors are historically
sourced, and many of them *mean something*: in GOST 14202-69, green is water,
red is fire suppression, yellow is gas.

```{r hero, eval = has_ggplot2, fig.height = 4.6, echo = FALSE}
# Hero: terrain elevation in the `steppe` camouflage ramp, with contour ruling.
# Sequential color + technical-drawing contours = the whole package in one image.
v <- expand.grid(row = seq_len(nrow(volcano)), col = seq_len(ncol(volcano)))
v$elev <- as.vector(volcano)

ggplot(v, aes(col, row, fill = elev)) +
  geom_raster() +
  geom_contour(aes(z = elev), colour = "grey15", linewidth = 0.15, bins = 12) +
  scale_fill_sovpal("steppe", discrete = FALSE, name = "Elevation") +
  coord_equal(expand = FALSE) +
  labs(
    title    = "sovpal",
    subtitle = "steppe (7K / 4BO / 6K camouflage) over terrain elevation"
  ) +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

## The palettes at a glance

```{r gallery, fig.height = 6}
show_all_palettes()
```

Six **archival** palettes are faithful to a source document or reference; three
**visualization-optimized** palettes (`_lines`, `_core`, `_cvd`, plus the warm
diverging variant) reuse those exact hex values, reordered or filtered for
contrast and colorblind safety. `sovpal_palettes()` lists them all:

```{r enumerate}
sovpal_palettes()
```

## Colors that mean something

The distinctive thing about the GOST palettes is *semantic* color: each name
maps to a fixed meaning. Because `sovpal()` returns a **named** vector, you can
map categories to their correct colors with `scale_*_manual()` — water is always
green, gas always yellow, regardless of plotting order.

```{r pipeline, eval = has_ggplot2, fig.height = 4}
set.seed(1969)
fluids <- names(sovpal("gost14202"))
flow <- expand.grid(year = 1965:1980, fluid = fluids)
flow$throughput <- round(abs(
  rep(c(120, 40, 70, 55, 25, 30, 90, 45), each = 16) +
    rep(seq(0, 30, length.out = 16), times = 8) +
    rnorm(nrow(flow), 0, 8)
))
flow$fluid <- factor(flow$fluid, levels = fluids)

ggplot(flow, aes(year, throughput, fill = fluid)) +
  geom_area(colour = "grey15", linewidth = 0.15) +
  scale_fill_manual(values = sovpal("gost14202")) +
  labs(
    title    = "Pipeline throughput by fluid category",
    subtitle = "GOST 14202-69 colors mapped to their meanings (green = water, red = fire main)",
    x = NULL, y = "Throughput", fill = "Fluid"
  )
```

## Two tiers: archival vs. visualization-optimized

Faithful palettes are not always plot-ready. In `gost14202`, the pale `gas`
yellow has a contrast ratio of only 1.97 against white, so it washes out as thin
lines or small points (every other stop, including the dark `fuel` brown at
9.32, reads fine). Rather than silently dropping it, sovpal ships an explicit
`gost14202_lines` palette — the choice is visible in your code.

```{r tier-before, eval = has_ggplot2, fig.height = 3.4}
pts <- data.frame(
  x = rep(seq_along(fluids), each = 12) + rnorm(12 * 8, 0, 0.12),
  y = rnorm(12 * 8),
  fluid = factor(rep(fluids, each = 12), levels = fluids)
)

ggplot(pts, aes(x, y, colour = fluid)) +
  geom_point(size = 2.6) +
  scale_color_sovpal("gost14202") +
  labs(title = "Full gost14202 (8 colors)",
       subtitle = "gas (yellow) barely reads as points on white",
       x = NULL, y = NULL)
```

```{r tier-after, eval = has_ggplot2, fig.height = 3.4}
lines_fluids <- names(sovpal("gost14202_lines"))
ggplot(subset(pts, fluid %in% lines_fluids), aes(x, y, colour = fluid)) +
  geom_point(size = 2.6) +
  scale_color_sovpal("gost14202_lines") +
  labs(title = "gost14202_lines (7 colors)",
       subtitle = "the same palette minus the one washed-out yellow",
       x = NULL, y = NULL)
```

## Sequential and diverging

`steppe` is a camouflage terrain ramp (sand → protective green → dark brown).
Its lightness is not strictly monotonic, so it reads as terrain more than as a
strict magnitude scale — which is exactly what the hero image uses it for.

For diverging scales the midpoint should be the *lightest* point. The
visualization-optimized diverging palettes use a light cream center; the
prettiest is `hazard_warm` (green → cream → red):

```{r diverging-warm, eval = has_ggplot2, fig.height = 4.4}
v$centered <- v$elev - mean(v$elev)
lim <- max(abs(v$centered))

ggplot(v, aes(col, row, fill = centered)) +
  geom_raster() +
  scale_fill_sovpal("hazard_warm", discrete = FALSE,
                    limits = c(-lim, lim), name = "Δ elev") +
  coord_equal(expand = FALSE) +
  labs(title = "hazard_warm", subtitle = "green / cream / red — light center, dark arms") +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

When red–green colorblindness matters, `hazard_cvd` (blue → cream → red) carries
the same safe-to-danger meaning with distinguishable hues:

```{r diverging-cvd, eval = has_ggplot2, fig.height = 4.4}
ggplot(v, aes(col, row, fill = centered)) +
  geom_raster() +
  scale_fill_sovpal("hazard_cvd", discrete = FALSE,
                    limits = c(-lim, lim), name = "Δ elev") +
  coord_equal(expand = FALSE) +
  labs(title = "hazard_cvd", subtitle = "colorblind-safe alternative") +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

The archival `hazard` (green / grey / red) keeps its faithful GOST values and is
intended as the historical record rather than the default plotting choice.

## Qualitative distributions

When the colors are just an attractive qualitative set, `soviet_military` gives
six muted, period-correct hues:

```{r densities, eval = has_ggplot2, fig.height = 3.8}
set.seed(42)
classes <- names(sovpal("soviet_military"))
dens <- data.frame(
  value = c(rnorm(200, 5), rnorm(200, 6.5), rnorm(200, 4),
            rnorm(200, 7), rnorm(200, 5.5), rnorm(200, 8)),
  class = factor(rep(classes, each = 200), levels = classes)
)

ggplot(dens, aes(value, fill = class)) +
  geom_density(alpha = 0.65, colour = "grey15", linewidth = 0.2) +
  scale_fill_sovpal("soviet_military") +
  labs(title = "soviet_military", x = NULL, y = "Density", fill = NULL)
```

## Usage

```{r usage}
# Retrieve a palette (named vector)
sovpal("gost14202")

# n colors, or interpolate a continuous ramp
sovpal("gost14202", n = 4)
length(sovpal("steppe", n = 100, type = "continuous"))

# Slice or subset with base-R indexing — no special arguments
sovpal("gost14202")[c("water", "fire", "air")]
```

`palette_info()` reports provenance, intended use, colorblindness caveats, and
**computed** white-background contrast (WCAG ratio; the 3:1 line is the non-text
threshold):

```{r info}
info <- palette_info("constructivist")
info$source
info$contrast_white
info$low_contrast_on_white   # cream and yellow fall below 3:1
```

ggplot2 scales are provided for both aesthetics: `scale_color_sovpal()` /
`scale_colour_sovpal()` and `scale_fill_sovpal()`, each with `discrete` and
`direction` arguments. Base R works directly, since `sovpal()` returns a plain
vector:

```{r base-r, fig.height = 3.2}
barplot(rep(1, 8), col = sovpal("gost14202"), border = "grey15",
        names.arg = names(sovpal("gost14202")), las = 2, cex.names = 0.7,
        axes = FALSE)
```

## Historical notes

**GOST 14202-69 — pipeline identification.** Adopted in 1969, it standardized
the color-coding of industrial pipelines across all Soviet enterprises: eight
base colors for fluid categories, referenced by paint card number. The system
remains in use in post-Soviet facilities under inherited technical regulations.

**GOST 12.4.026 — safety signage.** Four signal colors encoding hierarchical
urgency: red (danger, prohibition), yellow (caution), green (safe, evacuation),
blue (mandatory action, PPE).

**4BO Protective Green (`#51653F`).** The canonical Soviet vehicle exterior
color, adopted in the 1930s and used through WWII and the Cold War. Based on
Vallejo 70.609 and corroborated by AK Interactive measurements of surviving
vehicles. It appears in `soviet_military` and as `steppe`'s middle stop.

**Cockpit turquoise (`#5B8F8C`).** The grey-turquoise of Soviet aircraft
interiors from the 1940s–1970s (MiG-15 through Su-27), derived from AK Real
Colors RC206.

**Constructivism.** The 1920s–30s poster language of red, black, and white with
yellow and blue accents. `constructivist` is *evocative* of this corpus, not
standards-derived; its cream (`#E8DFC8`) recalls the aged paper ground and is
low-contrast on white — hence `constructivist_core`.
```