---
title: "Introduction to sovpal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sovpal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 3.6,
  dpi = 110
)
```

```{r setup}
library(sovpal)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_jpeg    <- requireNamespace("jpeg", quietly = TRUE)
```

```{r theme, include = FALSE}
# theme_linedraw resembles a ruled technical drawing. The grid is lightened so
# the fill and colour aesthetics dominate the plot.
if (has_ggplot2) {
  library(ggplot2)
  theme_sovpal <- function(base_size = 11) {
    theme_linedraw(base_size = base_size) +
      theme(
        panel.grid.major = element_line(linewidth = 0.2, colour = "grey80"),
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey30"),
        legend.key       = element_rect(colour = NA)
      )
  }
  theme_set(theme_sovpal())
}

# Draw an artwork thumbnail beside its sampled palette.
lum <- function(hex) {
  rgb <- grDevices::col2rgb(hex) / 255
  colSums(c(0.2126, 0.7152, 0.0722) * rgb)
}
art_plot <- function(img_path, pal_name, caption) {
  im   <- jpeg::readJPEG(img_path)
  cols <- sovpal(pal_name); n <- length(cols)
  layout(matrix(1:2, nrow = 1), widths = c(1, 1.25))
  par(mar = c(0.4, 0.4, 2, 0.4))
  plot.new(); plot.window(c(0, 1), c(0, 1)); rasterImage(im, 0, 0, 1, 1)
  title(caption, font.main = 2, cex.main = 1)
  par(mar = c(0.4, 0.4, 2, 0.4))
  plot.new(); plot.window(c(0, n), c(0, 1))
  rect(0:(n - 1), 0, 1:n, 1, col = cols, border = "grey25", lwd = 1.5)
  text((0:(n - 1)) + 0.5, 0.5, names(cols), srt = 90, cex = 0.8,
       col = ifelse(lum(cols) > 0.5, "grey10", "white"))
}
```

`sovpal` provides colour palettes from three Soviet sources: GOST industrial
standards, military paint references, and avant-garde artworks. Several palettes
are semantic. In GOST 14202-69, green denotes water, red fire suppression, and
yellow flammable gas.

```{r hero, eval = has_ggplot2, fig.height = 4.6, echo = FALSE}
# Terrain elevation rendered with the steppe sequential palette.
v <- expand.grid(row = seq_len(nrow(volcano)), col = seq_len(ncol(volcano)))
v$elev <- as.vector(volcano)

ggplot(v, aes(col, row, fill = elev)) +
  geom_raster() +
  geom_contour(aes(z = elev), colour = "grey15", linewidth = 0.15, bins = 12) +
  scale_fill_sovpal("steppe", discrete = FALSE, name = "Elevation") +
  coord_equal(expand = FALSE) +
  labs(title = "sovpal",
       subtitle = "steppe (7K / 4BO / 6K camouflage) over terrain elevation") +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Palettes by domain

Each palette has a thematic `domain` and a functional `type` (qualitative,
sequential, or diverging).

```{r enumerate}
sovpal_palettes()
```

```{r gallery, fig.height = 7}
show_all_palettes()
```

## Industrial

The GOST palettes are semantic. Because `sovpal()` returns a named vector,
categories can be mapped to fixed colours with `scale_*_manual()`. The
assignment is by name, so water is green and gas is yellow regardless of
plotting order.

```{r pipeline, eval = has_ggplot2, fig.height = 4}
set.seed(1969)
fluids <- names(sovpal("gost14202"))
flow <- expand.grid(year = 1965:1980, fluid = fluids)
flow$throughput <- round(abs(
  rep(c(120, 40, 70, 55, 25, 30, 90, 45), each = 16) +
    rep(seq(0, 30, length.out = 16), times = 8) +
    rnorm(nrow(flow), 0, 8)
))
flow$fluid <- factor(flow$fluid, levels = fluids)

ggplot(flow, aes(year, throughput, fill = fluid)) +
  geom_area(colour = "grey15", linewidth = 0.15) +
  scale_fill_manual(values = sovpal("gost14202")) +
  labs(title = "Pipeline throughput by fluid category",
       subtitle = "GOST 14202-69 fluid colours",
       x = NULL, y = "Throughput", fill = "Fluid")
```

## Military

`soviet_military` is a set of period vehicle and aircraft colours. `steppe`
(shown above) is a camouflage terrain ramp.

```{r military, eval = has_ggplot2, fig.height = 3.8}
set.seed(42)
classes <- names(sovpal("soviet_military"))
dens <- data.frame(
  value = c(rnorm(200, 5), rnorm(200, 6.5), rnorm(200, 4),
            rnorm(200, 7), rnorm(200, 5.5), rnorm(200, 8)),
  class = factor(rep(classes, each = 200), levels = classes)
)
ggplot(dens, aes(value, fill = class)) +
  geom_density(alpha = 0.65, colour = "grey15", linewidth = 0.2) +
  scale_fill_sovpal("soviet_military") +
  labs(title = "soviet_military", x = NULL, y = "Density", fill = NULL)
```

## Artistic

Each artistic palette is colour-sampled from a specific artwork and named for
the artist, following the convention of the MetBrewer package. The hex values
are approximate, sampled from reproductions. See `data-raw/sample_artwork.R`.

```{r art-lissitzky, eval = has_jpeg, fig.height = 2.6, echo = FALSE}
art_plot("artwork/lissitzky.jpg", "lissitzky",
         "El Lissitzky, Beat the Whites with the Red Wedge (1919)")
```

```{r art-popova, eval = has_jpeg, fig.height = 2.6, echo = FALSE}
art_plot("artwork/popova.jpg", "popova",
         "Liubov Popova, Architectonic Painting (1917)")
```

```{r art-malevich, eval = has_jpeg, fig.height = 2.6, echo = FALSE}
art_plot("artwork/malevich.jpg", "malevich",
         "Kazimir Malevich, Suprematist Composition (1916)")
```

`stepanova` is sampled from Varvara Stepanova's *Billiard Players* (1920), a
cubo-futurist work in charcoal, slate, brick, gold, and khaki. The work is under
copyright in the EU (Stepanova died in 1958), so the image is not bundled; the
palette is shown below.
[View the work at the Museo Thyssen-Bornemisza](https://www.museothyssen.org/en/collection/artists/stepanova-varvara/billiard-players).

```{r art-stepanova, fig.height = 1.7}
show_palette("stepanova", notes = FALSE)
```

## Sequential ramps

`rust`, `steel`, `cobalt`, and `field` are constructed for monotonic lightness
and are suitable for continuous scales. `steppe` is multi-hue and its lightness
is not monotonic.

```{r sequential, eval = has_ggplot2, fig.height = 2.8}
seqs <- c("rust", "steel", "cobalt", "field")
ramp <- do.call(rbind, lapply(seqs, function(p)
  data.frame(pal = p, x = 1:100, fill = sovpal(p, n = 100, type = "continuous"))))
ramp$pal <- factor(ramp$pal, levels = seqs)

ggplot(ramp, aes(x, 1, fill = fill)) +
  geom_raster() +
  scale_fill_identity() +
  facet_wrap(~pal, ncol = 1, strip.position = "left") +
  labs(x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        strip.text.y.left = element_text(angle = 0, face = "bold"),
        panel.grid = element_blank())
```

## Composite

The diverging palettes place the lightest colour at the midpoint. `hazard_warm`
(green, cream, red) and `hazard_cvd` (blue, cream, red) use a light cream
midpoint; `hazard_cvd` is intended for red-green colour vision deficiency.
`hazard` (green, grey, red) retains the GOST source colours.

```{r diverging-warm, eval = has_ggplot2, fig.height = 4.4}
v$centered <- v$elev - mean(v$elev)
lim <- max(abs(v$centered))
ggplot(v, aes(col, row, fill = centered)) +
  geom_raster() +
  scale_fill_sovpal("hazard_warm", discrete = FALSE,
                    limits = c(-lim, lim), name = "Delta elev") +
  coord_equal(expand = FALSE) +
  labs(title = "hazard_warm", subtitle = "green, cream, red") +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Usage

```{r usage}
# Retrieve a palette (named vector)
sovpal("malevich")

# n colours, or interpolate a continuous ramp
sovpal("gost14202", n = 4)
length(sovpal("steppe", n = 100, type = "continuous"))

# Subset with base-R indexing
sovpal("gost14202")[c("water", "fire", "air")]
```

`palette_info()` reports the domain, provenance, recommended use, colour vision
notes, and computed white-background contrast (WCAG ratio; 3:1 is the threshold
for graphical objects). The off-white ground in `lissitzky` falls below the
threshold and is listed in `low_contrast_on_white`.

```{r info}
info <- palette_info("lissitzky")
info$source
info$contrast_white
info$low_contrast_on_white
```

## Historical notes

GOST 14202-69 (pipeline identification), adopted in 1969, standardised the
colour-coding of industrial pipelines across Soviet enterprises: eight base
colours for fluid categories, referenced by paint card number.

GOST 12.4.026 (safety signage) defines four signal colours: red (danger),
yellow (caution), green (safe), and blue (mandatory action).

4BO Protective Green (`#51653F`) was the standard Soviet vehicle exterior colour
from the 1930s through the Cold War. The value is based on Vallejo 70.609. It
appears in `soviet_military` and as the middle stop of `steppe`.

The artistic palettes (`lissitzky`, `popova`, `stepanova`, `malevich`) are
colour-sampled from reproductions of Constructivist and Suprematist works made
between 1916 and 1920. The values are approximate and are not authoritative
colour records.
```