---
title: "Introduction to sovpal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sovpal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 3.6,
  dpi = 110
)
```

```{r setup}
library(sovpal)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_jpeg    <- requireNamespace("jpeg", quietly = TRUE)
```

```{r theme, include = FALSE}
# An opinionated visual language for the showcase. theme_linedraw evokes the
# ruled, boxed look of Soviet technical drawing (the GOST/ESKD drafting world
# these colors come from); we lighten its grid so the COLOR carries the plot.
if (has_ggplot2) {
  library(ggplot2)
  theme_sovpal <- function(base_size = 11) {
    theme_linedraw(base_size = base_size) +
      theme(
        panel.grid.major = element_line(linewidth = 0.2, colour = "grey80"),
        panel.grid.minor = element_blank(),
        plot.title       = element_text(face = "bold"),
        plot.subtitle    = element_text(colour = "grey30"),
        legend.key       = element_rect(colour = NA)
      )
  }
  theme_set(theme_sovpal())
}

# Draw an artwork thumbnail beside its sampled palette (MetBrewer style).
lum <- function(hex) {
  rgb <- grDevices::col2rgb(hex) / 255
  colSums(c(0.2126, 0.7152, 0.0722) * rgb)
}
art_plot <- function(img_path, pal_name, caption) {
  im   <- jpeg::readJPEG(img_path)
  cols <- sovpal(pal_name); n <- length(cols)
  layout(matrix(1:2, nrow = 1), widths = c(1, 1.25))
  par(mar = c(0.4, 0.4, 2, 0.4))
  plot.new(); plot.window(c(0, 1), c(0, 1)); rasterImage(im, 0, 0, 1, 1)
  title(caption, font.main = 2, cex.main = 1)
  par(mar = c(0.4, 0.4, 2, 0.4))
  plot.new(); plot.window(c(0, n), c(0, 1))
  rect(0:(n - 1), 0, 1:n, 1, col = cols, border = "grey25", lwd = 1.5)
  text((0:(n - 1)) + 0.5, 0.5, names(cols), srt = 90, cex = 0.8,
       col = ifelse(lum(cols) > 0.5, "grey10", "white"))
}
```

`sovpal` provides color palettes drawn from the Soviet world — GOST
pipeline-identification standards, military paint specifications, and the
palettes of avant-garde artworks. Colors are historically sourced, and many of
them *mean something*: in GOST 14202-69, green is water, red is fire
suppression, yellow is gas.

```{r hero, eval = has_ggplot2, fig.height = 4.6, echo = FALSE}
# Hero: terrain elevation in the `steppe` camouflage ramp, with contour ruling.
v <- expand.grid(row = seq_len(nrow(volcano)), col = seq_len(ncol(volcano)))
v$elev <- as.vector(volcano)

ggplot(v, aes(col, row, fill = elev)) +
  geom_raster() +
  geom_contour(aes(z = elev), colour = "grey15", linewidth = 0.15, bins = 12) +
  scale_fill_sovpal("steppe", discrete = FALSE, name = "Elevation") +
  coord_equal(expand = FALSE) +
  labs(title = "sovpal",
       subtitle = "steppe (7K / 4BO / 6K camouflage) over terrain elevation") +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Palettes by domain

Each palette has a thematic **domain** and a functional **type**
(qualitative / sequential / diverging):

```{r enumerate}
sovpal_palettes()
```

```{r gallery, fig.height = 7}
show_all_palettes()
```

## Industrial — colors that mean something

The distinctive thing about the GOST palettes is *semantic* color. Because
`sovpal()` returns a **named** vector, you can map categories to their correct
colors with `scale_*_manual()` — water is always green, gas always yellow,
regardless of plotting order.

```{r pipeline, eval = has_ggplot2, fig.height = 4}
set.seed(1969)
fluids <- names(sovpal("gost14202"))
flow <- expand.grid(year = 1965:1980, fluid = fluids)
flow$throughput <- round(abs(
  rep(c(120, 40, 70, 55, 25, 30, 90, 45), each = 16) +
    rep(seq(0, 30, length.out = 16), times = 8) +
    rnorm(nrow(flow), 0, 8)
))
flow$fluid <- factor(flow$fluid, levels = fluids)

ggplot(flow, aes(year, throughput, fill = fluid)) +
  geom_area(colour = "grey15", linewidth = 0.15) +
  scale_fill_manual(values = sovpal("gost14202")) +
  labs(title = "Pipeline throughput by fluid category",
       subtitle = "GOST 14202-69 colors mapped to their meanings",
       x = NULL, y = "Throughput", fill = "Fluid")
```

## Military — paint and camouflage

`soviet_military` is an earth-toned set of period vehicle and aircraft colors;
`steppe` (the hero image above) is the camouflage terrain ramp.

```{r military, eval = has_ggplot2, fig.height = 3.8}
set.seed(42)
classes <- names(sovpal("soviet_military"))
dens <- data.frame(
  value = c(rnorm(200, 5), rnorm(200, 6.5), rnorm(200, 4),
            rnorm(200, 7), rnorm(200, 5.5), rnorm(200, 8)),
  class = factor(rep(classes, each = 200), levels = classes)
)
ggplot(dens, aes(value, fill = class)) +
  geom_density(alpha = 0.65, colour = "grey15", linewidth = 0.2) +
  scale_fill_sovpal("soviet_military") +
  labs(title = "soviet_military", x = NULL, y = "Density", fill = NULL)
```

## Artistic — palettes from the avant-garde

Each artistic palette is **color-sampled from a specific artwork** and named for
the artist (cf. \pkg{MetBrewer}). The hex values are approximate, sampled from
reproductions — see `data-raw/sample_artwork.R`.

```{r art-lissitzky, eval = has_jpeg, fig.height = 2.6, echo = FALSE}
art_plot("artwork/lissitzky.jpg", "lissitzky",
         "El Lissitzky — Beat the Whites with the Red Wedge (1919)")
```

```{r art-popova, eval = has_jpeg, fig.height = 2.6, echo = FALSE}
art_plot("artwork/popova.jpg", "popova",
         "Liubov Popova — Architectonic Painting (1917)")
```

```{r art-malevich, eval = has_jpeg, fig.height = 2.6, echo = FALSE}
art_plot("artwork/malevich.jpg", "malevich",
         "Kazimir Malevich — Suprematist Composition (1916)")
```

`stepanova` is sampled from Varvara Stepanova's *Billiard Players* (1920) — a
muted cubo-futurist work of charcoal, slate, brick, gold, and khaki. Her work is
still under copyright in the EU (she died in 1958), so the image is not bundled;
[view it at the Museo Thyssen-Bornemisza](https://www.museothyssen.org/en/collection/artists/stepanova-varvara/billiard-players).

```{r art-stepanova, fig.height = 1.7}
show_palette("stepanova", notes = FALSE)
```

## Composite — diverging scales

For a diverging scale the midpoint should be the *lightest* point. The composite
diverging palettes use a light cream center; `hazard_warm` (green → cream → red)
is the most legible:

```{r diverging-warm, eval = has_ggplot2, fig.height = 4.4}
v$centered <- v$elev - mean(v$elev)
lim <- max(abs(v$centered))
ggplot(v, aes(col, row, fill = centered)) +
  geom_raster() +
  scale_fill_sovpal("hazard_warm", discrete = FALSE,
                    limits = c(-lim, lim), name = "Δ elev") +
  coord_equal(expand = FALSE) +
  labs(title = "hazard_warm", subtitle = "green / cream / red") +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        axis.ticks = element_blank())
```

When red–green colorblindness matters, `hazard_cvd` (blue → cream → red) carries
the same safe-to-danger meaning with distinguishable hues. The archival `hazard`
(green / grey / red) keeps its faithful GOST values as the historical record.

## Usage

```{r usage}
# Retrieve a palette (named vector)
sovpal("malevich")

# n colors, or interpolate a continuous ramp
sovpal("gost14202", n = 4)
length(sovpal("steppe", n = 100, type = "continuous"))

# Slice or subset with base-R indexing
sovpal("gost14202")[c("water", "fire", "air")]
```

`palette_info()` reports domain, provenance, intended use, colorblindness
caveats, and **computed** white-background contrast (WCAG ratio; 3:1 is the
non-text threshold). Lissitzky's off-white ground is flagged automatically:

```{r info}
info <- palette_info("lissitzky")
info$source
info$contrast_white
info$low_contrast_on_white
```

## Historical notes

**GOST 14202-69 — pipeline identification.** Adopted in 1969, it standardized
the color-coding of industrial pipelines across all Soviet enterprises: eight
base colors for fluid categories, referenced by paint card number.

**GOST 12.4.026 — safety signage.** Four signal colors encoding urgency: red
(danger), yellow (caution), green (safe), blue (mandatory action).

**4BO Protective Green (`#51653F`).** The canonical Soviet vehicle exterior
color, used from the 1930s through the Cold War; based on Vallejo 70.609. It
appears in `soviet_military` and as `steppe`'s middle stop.

**The avant-garde palettes.** `lissitzky`, `popova`, `stepanova`, and `malevich`
are color-sampled from reproductions of specific Constructivist and Suprematist
works (1916–1920). They are evocative of those works rather than authoritative
color records.
```